\large{Vipassana for Hackers} \\ \Huge{The Proposal} \\ \large\textit{Version 0.1}

### 2019-12-04 the-proposal/the-proposal.tex
proposal: quantitative case study description

Qualitative research into the specific nature of human consciousness or the broader
nature of consciousness as it applies to any organism or artificial intelligence can
only be performed directly. In this case, a case study performed by an individual
(myself) will explore the question of consciousness and the mind-body problem
directly, through Vipassana meditation, for a prolonged period of time (the remainder
of my life). Many theories and questions currently exist regarding
consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism}, and
the \textit{Simulation Hypothesis}. This study will not address any one of them
directly, nor will it absorb any specific hypothesis. The data recovered from
practicing Vipassana meditation directly for prolonged periods of time will almost
certainly overlap with many such ideas but themes and narratives are much more likely
to be emergent, rather than conforming to an existing hypothesis of consciousness.

@@

Vipassana's instruction provides us with claims we can evaluate: ``[Vipassana] explores the entire field of
mind and matter'' \cite{goenka1999discourses}, ``[Vipassana] is a technique that will
eradicate suffering'' \cite{dhammaorg2014code}, and ``Vipassana aims at ... total
liberation and full enlightenment'' \cite{dhammaorg2014code}.

>>>>

Qualitative research into the specific nature of human consciousness or the broader
nature of consciousness as it applies to any organism or artificial intelligence can
only be performed directly. In this case, a case study performed by an individual
(myself) will explore the question of consciousness and the mind-body problem
directly, through Vipassana meditation, for a prolonged period of time (the remainder
of my life). Many theories and questions currently exist regarding
consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism},
\textit{Theory of the Entropic Brain},
the \textit{Simulation Hypothesis}, etc. This study will not address any one of the
many theories of consciousness
directly, nor will it absorb any specific hypothesis. The data recovered from
practicing Vipassana meditation directly for prolonged periods of time will almost
certainly overlap with many such ideas but themes and narratives are likely
to be emergent, rather than conforming to an existing hypothesis of consciousness.

@@

Vipassana's instruction provides us with claims we can evaluate: ``[Vipassana] explores the entire field of
mind and matter'' \cite{goenka1999discourses}, ``[Vipassana] is a technique that will
eradicate suffering'' \cite{dhammaorg2014code}, and ``Vipassana aims at ... total
liberation and full enlightenment'' \cite{dhammaorg2014code}. Progress made in
evaluation of these claims will be, by its very nature, emergent.

The validity of these emergent observations made during the qualitative portion of
the study can be asserted across most axes of validity: prolonged engagement, rich
descriptions, external audits (from more experienced meditators), identification of
researcher bias, peer debriefing (again from more experienced meditators), and
searching for discrepant evidence are all possible. Respondent validation (member
checking), although possible, may not carry much weight in terms of strengthening the
validity of the study, given an initial sample of one. Triangulation will not be
possible with a sample of one. If the study proves productive, in the future
additional researcher-subjects intent on a lifelong Vipassana practice may strengthen
respondent validation and triangulation for parallel studies. Triangulation of
qualitative analysis of consciousness suffers from the paradox that any individual's
consciousness is only directly observable by that individual and conclusions --- even
those validated by triangulation --- will always be in the third-person.

Data will be collected on a daily basis and themes and narratives regularly collected
with the intention of
describing mental phenomena and the evolution of traits as Vipassana practice
progresses. Monthly or yearly schedules will be decided with a
supervisor but an example daily schedule (subject to change before the study begins)
is available in \textit{Appendix 1: Proposed Daily Schedule}.
